Identifying Impact of Vitamin C Treatments on Tooth Growth in Guinea Pigs
============================================================================
Author: Lowell Ricketts  
Date: March 2016

# Synopsis

Using hypothesis testing and associated confidence intervals, I find that 
delivering vitamin C supplements in the form of orange juice rather than ascorbic 
acid leads to greater length of odontoblasts (cells responsible for tooth growth) 
when the dosage levels of the supplement are less than 2 mg/day. At levels of 2 
mg/day there was no significant difference in odontoblast length between the two 
groups.

# Introduction

This project was completed during the Statistical Inference course offered by 
Coursera in the Data Science specialization tract. The course was taught by Brian
Caffo from Johns Hopkins Bloomberg School of Public Health.

For this project I use the [ToothGrowth dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html) 
found in the R Datasets package. The data measures the response in the length
of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each
animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by
one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C.)

In the following analysis I begin with an exploratory look at the underlying 
data. Next, I conduct hypothesis tests to see if there is a statistically 
significant difference between mean tooth length for the two delivery methods.

Please see the Appendix for all supporting R code.

# Exploratory Analysis

Summary statistics for tooth length are reported in the table below
by supplement type and dosage. Standard deviations suggest that tooth length
is more widely dispersed for guinea pigs receiving OJ than VC at lower dosages, 
while VC has the highest variability at the highest dose. When comparing the two 
supplements by dosage, mean tooth length is greater for OJ across low dosages 
and about equal at the highest dose.

```{r, echo = FALSE}
library(datasets)
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))

# Load ToothGrowth data: The Effect of Vitamin C on Tooth Growth in Guinea Pigs
# Variable List:
# len = Tooth length
# supp = Supplement type (VC = ascorbic acid, OJ = orange juice)
# dose = Dose in milligrams/day
data(ToothGrowth)

# Calculate summary statistics for tooth length, by supplement and dosage
summarize(group_by(ToothGrowth, supp, dose), mean = mean(len), sd = sd(len),
          min = min(len), max = max(len))
```


Next, Figure 1 shows the distribution of tooth length by supplement and dosage.
The distributions look gaussian with a high variance. Given the limited sample
size I assume that the t-distribution is best suited to draw statistical 
inferences. 

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
# Plot the distribution of tooth length, by supplement and dosage
# It appears that OJ has greater tooth lengths across lower dosages, and similar
# lengths at the highest dosage
ggplot(ToothGrowth, aes(len)) + 
        geom_density(fill = "lightblue") + 
        facet_wrap(~supp + dose) + 
        ggtitle("Figure 1: Tooth Length Distribution, By Supplement & Dosage") +
        labs(x = "Tooth Length", y = "Density")
```

# Hypothesis Testing

For the t-tests I assume that the observations are not paired given that all 60
guinea pigs were only sampled once. I don't assume that the groups have equal 
variance, the results of the following t-tests vary little based on that 
assumption. I define my type I error rate as 5 percent and use 95 percent 
confidence intervals for the estimated difference in group means.

The first test is whether the mean tooth length of the group given orange juice
differs from the mean tooth length of the group given ascorbic acid. For this
initial t-test I use all observations, regardless of dosage.

```{r, echo = FALSE}
t.test(ToothGrowth$len ~ ToothGrowth$supp, paired = FALSE)
```

The null hypothesis, that the difference in means is equal to zero, cannot be
rejected at my designated significance level given that the p-value is greater 
than 0.05. This is also reflected in the confidence interval, zero is present on 
the lower end of the range. Certainly, if the confidence interval was narrowed 
by using a 10 percent type I error rate then we could say that OJ is more 
effective at effecting tooth growth. However, its important to stick to our 
significance thresholds rather than change them to tell a different story.

Next, I conduct a t-test for each of the three dosage levels to see if the 
delivery methods have different effectiveness based on dosage. Figure 2 shows 
the 95 percent confidence intervals for the difference in means across dosage 
levels. These confidence intervals were calculated using the same t-tests, see
the Appendix for the exact R code used for the tests. The estimated difference
in means from is shown in the center of the intervals.

The difference between the group means is positive and statistically significant
for both the 0.5 and 1 mg/day dosage levels. Thus, we can conclude that orange
juice leads to greater tooth growth when dosage levels are lower than 2 mg/day.
The confidence interval for 2 mg/day shows that there is no significant
difference in tooth growth between groups. In terms of our null 
hypothesis' of a zero difference in group means, we reject the null for 0.5 and 
1 mg/day dosages and cannot reject the null at the 2 mg/day dosage.

```{r, echo = FALSE, fig.width = 7, fig.height = 5}
# Low Dose: Difference in mean highly significiant
lowdose <- filter(ToothGrowth, dose == 0.5)
low <- t.test(lowdose$len ~ lowdose$supp, paired = FALSE)
ci1 <- low$conf.int[1:2]
mu1 <- low$estimate[1] - low$estimate[2]

# Moderate Dose: Difference in mean highly significiant
meddose <- filter(ToothGrowth, dose == 1)
med <- t.test(meddose$len ~ meddose$supp, paired = FALSE)
ci2 <- med$conf.int[1:2]
mu2 <- med$estimate[1] - med$estimate[2]

# High Dose: Null is not rejected, CI is very wide
highdose <- filter(ToothGrowth, dose == 2)
high <- t.test(highdose$len ~ highdose$supp, paired = FALSE)
ci3 <- high$conf.int[1:2]
mu3 <- high$estimate[1] - high$estimate[2]

ci_intervals <- cbind(rbind(ci1, ci2, ci3),
                           dose = c(0.5, 1, 2),
                        mean = c(mu1, mu2, mu3))
ci_intervals <- data.frame(ci_intervals)
ci_intervals <- rename(ci_intervals, lower = V1, higher = V2)

g <- ggplot(ci_intervals, aes(x = dose, y = mean)) +
        geom_point(size = 4) +
        geom_errorbar(aes(ymax = higher, ymin = lower)) +
        geom_hline(yintercept = 0, size = 1) +
        labs(x = "Dose", y = "95% Confidence Interval") +
        ggtitle("Figure 2: Difference Between OJ and VC Group Means, By Dose")
g
```

# Conclusion

Orange juice is a more effective way to increase tooth length in guinea pigs 
when treatment dosages are lower than 2 mg/day. At dosages of 2 mg/day neither
method (orange juice or ascorbic acid) was significantly more effective than the
other.

# Appendix

```{r, eval = FALSE}

library(datasets)
library(dplyr)
library(ggplot2)

# Load ToothGrowth data: The Effect of Vitamin C on Tooth Growth in Guinea Pigs
# Variable List:
# len = Tooth length
# supp = Supplement type (VC = ascorbic acid, OJ = orange juice)
# dose = Dose in milligrams/day
data(ToothGrowth)
head(ToothGrowth)

# Calculate summary statistics for tooth length, by supplement and dosage
summarize(group_by(ToothGrowth, supp, dose), mean = mean(len), sd = sd(len),
          min = min(len), max = max(len))

# Plot the distribution of tooth length, by supplement and dosage
# It appears that OJ has greater tooth lengths across lower dosages, and similar
# lengths at the highest dosage
ggplot(ToothGrowth, aes(len)) + 
        geom_density(fill = "lightblue") + 
        facet_wrap(~supp + dose) + 
        ggtitle("Figure 1: Tooth Length Distribution, By Supplement & Dosage") +
        labs(x = "Tooth Length", y = "Density")

# Test difference in supplement type, was orange juice more effective than
# ascorbic acid? H0 = There was no difference in mean tooth length between VC 
# and OJ.
mu_oj <- mean(ToothGrowth$len[ToothGrowth$supp == "OJ"])
mu_vc <- mean(ToothGrowth$len[ToothGrowth$supp == "VC"])
sd_oj <- sd(ToothGrowth$len[ToothGrowth$supp == "OJ"])
sd_vc <- sd(ToothGrowth$len[ToothGrowth$supp == "VC"])
n_oj <- length(ToothGrowth$len[ToothGrowth$supp == "OJ"])
n_vc <- length(ToothGrowth$len[ToothGrowth$supp == "VC"])

# Pooled variance estimate
sp <- sqrt( ((29 * (sd_oj ^ 2)) + 29 * (sd_vc ^ 2)) / 
        (n_oj + n_vc -2))

# 95 % confidence interval for the difference in supplement group means
(mu_oj - mu_vc) + c(-1,1) * qt(.975, 58) * sp * (1/30 + 1/30) ^ .5

# Using a 90% confidence interval
(mu_oj - mu_vc) + c(-1,1) * qt(.95, 58) * sp * (1/30 + 1/30) ^ .5

t.test(ToothGrowth$len ~ ToothGrowth$supp, paired = FALSE)

# Does the difference between OJ and VC change by dose?
# Low Dose: Difference in mean highly significiant
lowdose <- filter(ToothGrowth, dose == 0.5)
t.test(lowdose$len ~ lowdose$supp, paired = FALSE)
low <- t.test(lowdose$len ~ lowdose$supp, paired = FALSE)
low
c1 <- low$conf.int[1:2]
mu1 <- low$estimate[1] - low$estimate[2]

# Moderate Dose: Difference in mean highly significiant
meddose <- filter(ToothGrowth, dose == 1)
t.test(meddose$len ~ meddose$supp, paired = FALSE)
med <- t.test(meddose$len ~ meddose$supp, paired = FALSE)
med
ci2 <- med$conf.int[1:2]
mu2 <- med$estimate[1] - med$estimate[2]

# High Dose: Null is not rejected, CI is very wide
highdose <- filter(ToothGrowth, dose == 2)
t.test(highdose$len ~ highdose$supp, paired = FALSE)
high <- t.test(highdose$len ~ highdose$supp, paired = FALSE)
high
ci3 <- high$conf.int[1:2]
mu3 <- high$estimate[1] - high$estimate[2]

ci_intervals <- cbind(rbind(ci1, ci2, ci3),
                           dose = c(0.5, 1, 2),
                        mean = c(mu1, mu2, mu3))
ci_intervals <- data.frame(ci_intervals)
ci_intervals <- rename(ci_intervals, lower = V1, higher = V2)

g <- ggplot(ci_intervals, aes(x = dose, y = mean)) +
        geom_point(size = 4) +
        geom_errorbar(aes(ymax = higher, ymin = lower)) +
        geom_hline(yintercept = 0, size = 1) +
        labs(x = "Dose", y = "95% Confidence Interval") +
        ggtitle("Figure 2: Difference Between OJ and VC Group Means, By Dose")
g
```
